PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes

### 2022-09-08 manuscript_00.Rmd
text updates

Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries impossible. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].

@@

The aggregateAcrossCells function from the scuttle R package was used to aggregate gene counts across species, study accession, and category where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017]. If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test. Any genes with a sum of 0 counts after aggregation were removed. The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + category". Contrasts were extracted either with the target category (e.g. "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a]. 
<!-- Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.  -->

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.91 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI's integration parameters were chosen rigorously by testing inegration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021]. Each study in the resource were hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space. The small number of outlier studies were removed from the resource.

After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`). The scEiaD database contains `r num2` non-ocular human and mouse tissues and `r num1` ocular tissues across human, chicken, macaque, and mouse (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`). This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`). To recognize these papers, we prominently feature each of these studies along with a direct link to their citation, where possible, on the loading page of plae.nei.nih.gov.

@@

timing_plot <- timings %>% as_tibble(rownames = 'Test') %>% 
  pivot_longer(-Test) %>% 
  mutate(Test = factor(Test, levels = row.names(timings))) %>% 
  mutate(Text = case_when(is.na(value) ~ 'Not possible',
                          TRUE ~ '')) %>% 
  ggplot(aes(x=name,y=value, label = Text)) + 
  geom_bar(stat= 'identity') + 
  geom_text(aes(y=0), hjust = 0, angle = 90) +
  facet_wrap(~Test, scales = 'free_y', ncol = 2) + 
  xlab('Resource') + 
  ylab('Seconds to Load') + 
  cowplot::theme_cowplot() +
  scale_y_continuous(breaks = breaks_fun)

@@

The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence in a consistent manner, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.

@@

While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently[@voigtSpectacleInteractiveResource2020]. PLAE is built around the scEiaD database, which use a multi-stage pipeline to build and identify a high performing scVI-based single cell RNA eye model that is used to integrate all datasets together [@swamyBuildingMegaSinglecell2021]. This allows for gene queries to be constructed and analyzed across studies. In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies. Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type). Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which can make local (on personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26). Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d). The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.

@@

PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values. So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells. The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells. Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).

For users interested in seeing how expression of a gene relates to study or other covariates, the Expression Plot view displays expression by cell type, predicted cell type, or cluster. The values can then be split and colored by multiple covariates like study or organism. This allows a user to see whether a gene pattern is consistent across studies or species. We also offer a Dot Plot view which can display a large number of genes in a space efficient manner. This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).

@@

```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app. A. UMAP view showing CRX expression with cell type labels, B. Information table showing expression of Crx across celltypes and organism. C. Box plot of selected genes in a user selected cell types across different studies. D. In silico in-situ of Rho. E. Dotplot demonstating efficient display of gene expression across many cell types.')
knitr::include_graphics('fig/fig1.png')
```

@@

Cell type marker genes can be proposed on the basis of existing knowledge about their function and on high expression differences when comparing to other cell types. We can further narrow down the list of markers by using the high diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers. We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`). These cell types were then assessed to identify differentially expressed genes which met the following criteria across human, mouse, and macaque: 1. padj less than $1 x 10^-4$
in two or more species, 2. mean log2 fold change greater than 2, and 3. mean padj less than $1 x 10^-5$. A final filter was applied to remove genes as candidates if they were differentially expressed (using the criteria above) in more than three different cell types. This left us with `r top_genes %>% nrow()` gene - cell type markers, a substantial reduction from the `r diff_tab %>% nrow()` we had initially. We provide a table of these genes as supplementary file "consistently_differentially_expressed_genes.csv.gz."

To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`). We see that human, mouse, and macaque have similar expression patterning of the proposed markers across the well supported cell types. The cell types are arranged by the columns by how related the expression patterns are. We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are near the RPC from which they derive from. The photoreceptors are also next to each other.

@@

Genes with few citations can be considered as novel markers. We see, for example, Onecut3 appears to well separate the horizontal cells and has few PubMed citations. To assess whether there are any other good candidates with few citations, we produced another heatmap (`r supFig_cap(name = 'undersupported_genes', display = 'cite')`). Most of these genes are differentially expressed across more than one cell type. We hand selected C1ql2 and Cartpt (amacrine), Frzb and Slitrk2 (Muller glia), and Onecut3 (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`). All five look to be specific to a cell type, except for Frzb which also is expressed in the RPE.

Other possibilities for ranking genes are possible. We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how one can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization. The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.

@@

```{r consist_genes_heatmap, fig.height=22, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms. Rows are split by what cell type the gene is a marker for. Columns are split by the cell types. More red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative. Relative number of citations are shown for the gene or gene " and retina" when searching against PubMed compared to 1000 randomly chosen genes.')

@@

To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types. We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`). To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types. We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C). We find one gene, ZBTB38, that meets this criteria and drops in expression when comparing RPC to neurogenic. We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C). We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC. Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.

@@

While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed. To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html". In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner. This left us with 26678 cells across six studies. Yan et al. previously proposed Prph2 and Rs1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al. proposed Calb1 and Gngt1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].

Our differential testing found that while Prph2, Rs1, and Gngt1 had expression differences matching that the findings Yan et al. and Peng et al. published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three. In contrast Calb1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87. We find six more genes with a padj < 0.01: Pde6g, Akap12, Hspb1, Gadd45b, Cc2d2a, and RP1l1 (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`). None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).

@@

The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and most ocular cell types. Earlier we proposed well supported cell types. Right now, most of these well supported cell types are in the retina. The RPE behind the retina, the cornea and lens in front, and the outflow tract and supporting musculature still could use more independent studies. Many common human diseases underlie the RPE (e.g. AMD) and the cornea (e.g. Keratoconus) and these tissues will remain a focus for the next update of our database. While we have curated a large number of cell types across the eye, many of these cell types have "sub" cell types. For examplethe human has three types of cones that are optimized for short, medium, and long wavelengths. More dramatically, Yan et al. have proposed over sixty types of mouse amacrine cell types [@yanMouseRetinalCell2020a]. While we have found it fairly straightforward to use machine learning to transfer the broad (e.g. rod, cone, amacrine) cell type labels across all datasets, we have found it very challenging to transfer the "sub" cell type labels. This is largely because while we have high diversity of studies for the broad cell types, this is not true for most of the sub celltypes. We believe more datasets that carefully dissect the sub-celltypes may be necessary to confidently propose the sub-celltypes of the retina. Finally, while our scVI based model can theoretically be used by an outside group to transfer their internal data into the scEiaD latent space, it is very challenging to implement as the the software landscape is still rapidly changing and it is very difficult to architect consistent outputs over time with regularly changing software versions. We are investigating simpler methods to enable transfer of knowledge from scEiaD to outside data.

@@

We have improved on the scEiaD dataset we published in Swamy et al. [@swamyBuildingMegaSinglecell2021] by adding another species (chicken), new ocular tissues, a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets. We re-structured the data into a custom database to enable our custom built high performance web app to quickly access the data. This democratizes access to over a million cells across `r meta_filter$study_accession %>% unique() %>% length()` studies to anyone with a web browser. We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying plae in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species. We provide a short example of how a custom query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.

>>>>

Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device. Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021]. Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging. In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020]. However each data deposit is independently processed, which makes cross study queries impossible. We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021].

@@

The aggregateAcrossCells function from the scuttle R package was used to aggregate gene counts across species, study accession, and category where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017]. If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test. Any genes with a sum of 0 counts after aggregation were removed. The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + category". Contrasts were extracted either with the target category (e.g. "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a].

<!-- Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.  -->

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021]. We updated the scEiaD database for the PLAE v0.91 web app in six important ways. First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021]. Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020]. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020]. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20]. As before, the pipeline's scVI's integration parameters were chosen rigorously by testing inegration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021]. Each study in the resource was hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space and a small number of outlier studies were removed from the resource.

After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`). The scEiaD database contains `r num2` non-ocular human and mouse tissues and `r num1` ocular tissues across human, chicken, macaque, and mouse (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`). This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`). To recognize these papers, we prominently feature each of these studies along with a direct link to their citation, where possible, on the loading page of plae.nei.nih.gov.

@@

timing_plot <- timings %>% as_tibble(rownames = 'Test') %>% 
  pivot_longer(-Test) %>% 
  mutate(Test = factor(Test, levels = row.names(timings))) %>% 
  mutate(Text = case_when(is.na(value) ~ 'Not\npossible',
                          TRUE ~ '')) %>% 
  ggplot(aes(x=name,y=value, label = Text)) + 
  geom_bar(stat= 'identity') + 
  geom_text(aes(y=0), hjust = 0, angle = 90) +
  facet_wrap(~Test, scales = 'free_y', ncol = 2) + 
  xlab('Resource') + 
  ylab('Seconds to Load') + 
  cowplot::theme_cowplot() +
  scale_y_continuous(breaks = breaks_fun)

@@

The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app. PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence in a consistent manner, integrating the independent studies to deliver cross study comparisons, and, crucially, providing full availability of the underlying data for outside usage.

@@

While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently[@voigtSpectacleInteractiveResource2020]. PLAE is built around the scEiaD database, which use a multi-stage pipeline to build and identify a high performing scVI-based single cell RNA eye model that is used to integrate all datasets together [@swamyBuildingMegaSinglecell2021]. This allows for gene queries to be constructed and analyzed across studies. In contrast, Spectacle, as of August 2022, is a compilation of study-level datasets and which makes queries across studies impossible. Spectacle does not contain any non-ocular datasets and the cell type labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type). Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.

@@

There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage. The major downside is that this approach requires a large amount of disk storage, which can make local (on a personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB. Second, the visualization code in PLAE (e.g. the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time. In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26). Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d). The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.

@@

PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values. So thus, a user can see *Crx* expression across over a million cells and separately see cell type labels for those same million cells. The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells. Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).

For users interested in seeing how expression of a gene relates to study or other covariates, the Expression Plot view displays expression by cell type, predicted cell type, or cluster. The values can then be split and colored by multiple covariates like study or organism. This allows a user to see whether a gene pattern is consistent across studies or species. We also offer a Dot Plot view which can display a large number of genes in a space efficient manner. This allows users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).

@@

```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app. A. UMAP view showing *Crx* expression with cell type labels, B. Information table showing expression of Crx across celltypes and organism. C. Box plot of selected genes in a user selected cell types across different studies. D. In silico in-situ of Rho. E. Dotplot demonstating efficient display of gene expression across many cell types.')
knitr::include_graphics('fig/fig1.png')
```

@@

Cell type marker genes can be proposed on the basis of existing knowledge about their function and on high expression differences when comparing to other cell types. We can further narrow down the list of markers by using the high diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers. We first identify a set of well supported cell types which we define as cell types that are detected in two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`). These cell types were then assessed to identify differentially expressed genes which met the following criteria across human, mouse, and macaque: 1. padj less than $1 x 10^-4$
in two or more species, 2. mean log2 fold change greater than 2, and 3. mean padj less than $1 x 10^-5$. A final filter was applied to remove genes as candidates if they were differentially expressed (using the criteria above) in more than three different cell types. This left us with `r top_genes %>% nrow()` gene - cell type markers, a substantial reduction from the `r diff_tab %>% nrow()` we had initially. We provide a table of these genes as supplementary file "consistently_differentially_expressed_genes.csv.gz."

To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`). We see that human, mouse, and macaque have similar expression patterning of the proposed markers across the well supported cell types. The cell types are arranged by the columns by how related the expression patterns are. We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are near the RPC from which they are derived.

@@

Genes with few citations can be considered as novel markers. We see, for example, *Onecut3* appears to well separate the horizontal cells and has few PubMed citations. To assess whether there are any other good candidates with few citations, we produced another heatmap (`r supFig_cap(name = 'undersupported_genes', display = 'cite')`). Most of these genes are differentially expressed across more than one cell type. We hand selected *C1ql2* and *Cartpt* (amacrine), *Frzb* and *Slitrk2* (Muller glia), and *Onecut3* (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`). All five look to be specific to a cell type, except for *Frzb* which also is expressed in the RPE.

Other strategies for ranking genes are possible. We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how one can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization. The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across many cells and studies.

@@

```{r consist_genes_heatmap, fig.height=22, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms. Rows are split by what cell type the gene is a marker for. Columns are split by the cell types. More intense red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative. Relative number of citations are shown for the gene or gene " and retina" when searching against PubMed compared to 1000 randomly chosen genes.')

@@

To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types. We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`). To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types. We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C). We find one gene, *Zbtb38*, that meets this criteria and drops in expression when comparing RPC to neurogenic. We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C). We replicate *`r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()`* from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC. Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.

@@

While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed. To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html". In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner. This left us with 26678 cells across six studies. Yan et al. previously proposed *Prph2* and *Rs1* as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al. proposed *Calb1* and *Gngt1* as cone region markers in macaque [@pengMolecularClassificationComparative2019a].

Our differential testing found that while *Prph2*, *Rs1*, and *Gngt1* had expression differences matching that the findings Yan et al. and Peng et al. published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three. In contrast *Calb1* did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87. We find six more genes with a padj < 0.01: *Pde6g*, *Akap12*, *Hspb1*, *Gadd45b*, *Cc2d2a*, and *RP1l1* (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`). None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).

@@

The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and most ocular cell types. Earlier we proposed well supported cell types. Right now, most of these well supported cell types are in the retina. The RPE behind the retina, the cornea and lens in front, and the outflow tract and supporting musculature still could use more independent studies. Many common human diseases underlie the RPE (e.g. AMD) and the cornea (e.g. Keratoconus) and these tissues will remain a focus for the next update of our database. While we have curated a large number of cell types across the eye, many of these cell types have "sub" cell types. For examplethe human has three types of cones that are optimized for short, medium, and long wavelengths. More dramatically, Yan et al. have proposed over sixty different mouse amacrine cell types [@yanMouseRetinalCell2020a]. While we have found it fairly straightforward to use machine learning to transfer the broad (e.g. rod, cone, amacrine) cell type labels across all datasets, we have found it very challenging to transfer the "sub" cell type labels. This is largely because while we have high diversity of studies for the broad cell types, this is not true for most of the sub celltypes. We believe more datasets that carefully dissect the sub-celltypes may be necessary to confidently propose the sub-celltypes of the retina. Finally, while our scVI based model can theoretically be used by an outside group to transfer their internal data into the scEiaD latent space, it is very challenging to implement as the the software landscape is still rapidly changing and it is very difficult to architect consistent outputs over time with regularly changing software versions. We are investigating simpler methods to enable transfer of knowledge from scEiaD to outside data.

@@

We have improved on the scEiaD dataset we published in Swamy et al. [@swamyBuildingMegaSinglecell2021] by adding another species (chicken), new ocular tissues, a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets. We re-structured the data into a custom database to enable our custom built high performance web app to quickly access the data. This democratizes access to over a million cells across `r meta_filter$study_accession %>% unique() %>% length()` studies to anyone with a web browser. We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying PLAE in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species. We provide a short example of how a novel query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.
